How to Avoid OOTA Without Really Trying

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Compress Section 5.2 ("Instruction Ordering")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Using this scheme, we say that an instruction $J$ is dependent on
another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
perhaps via intermediate instructions.
Any hardware dependency ultimately emanates from register initial values,
load instructions, immediate values, or I/O accesses.
We set aside I/O, so that load instructions are the only sources
of truly new information in a thread.
Dependencies are important because they affect instruction ordering in
weakly ordered architectures.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction or its outputs until the relevant inputs' sources have committed.
The reason is obvious: An input is subject to change at any time until
its source commits to its value,
and the instruction and its outputs can't commit until the CPU has fully
determined what it will do and what they will be.
(On the other hand, the CPU need not wait for inputs that won't affect
the instruction's results.
For instance, a conditional-move instruction wouldn't need to wait for
the source register input to be determined once it knew that the
condition was definitely false.
And, although no architectures we are aware of do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)

Of course, an instruction's inputs don't all have to be outputs from
earlier instructions; some of them may be immediate constants present
in the instruction itself.
In this case there is no need to wait for those inputs because
their values are fully determined from the start.

The overall effect of these hardware dependencies is that if a change to an
output of instruction $I$ would lead to a change in the action or outputs of
a later instruction $J$, then the CPU must commit $I$'s output before
committing $J$'s action and outputs.
Thus dependencies force instructions to commit in order, even on
weakly ordered architectures.

We will say that one instruction is \emph{ordered after} another to mean that
the CPU forces it to commit after the other one commits.
Hence instructions are ordered after the instructions they depend on.
(Note that this implies nothing about when a load instruction retrieves
its value from memory; it may do so long before it or a prior
instruction commits.)

Dependencies aren't the only mechanism that can lead to
ordering in hardware executions.
The simplest alternative is a load reading from a store in another
thread, i.e., the rfe relation.
As mentioned in Section~\ref{sec:Brief OOTA Overview}, a CPU does
not make the value of a store instruction available for other CPUs to
load until the store commits or some time later.
And after this happens, it takes some time for the information about
the store to travel from that CPU to others, owing to the inescapable
facts that processors have nonzero size and information cannot travel
faster than the speed of light.
Since the load instruction cannot commit until its output (the value it
reads) is fully determined, it must commit after the store it reads
from.
(To be fair, this should be considered more as a \emph{result} of
ordering than as a \emph{cause} of ordering, in that
if a load commits before a store on another thread then it
cannot possibly read from that store.)

A similarly straightforward mechanism for ordering is conditional or
computed branches.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to (that is, the output value it will
send to the instruction-pointer register);
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional- or computed-branch
instruction in a hardware execution must commit after the branch, and hence
after the source for the branch's condition or destination input.
(There could theoretically be an exception for a branch that
conditionally jumps to the immediately following instruction.
We can ignore this possibility by treating such branches as no-ops.)

>>>>

Using this scheme, we say that an instruction $J$ is dependent on
another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
perhaps via intermediate instructions.
Any hardware dependency ultimately emanates from register initial values,
load instructions, immediate values, or I/O accesses.
We set aside I/O, so that load instructions are the only sources
of truly new information in a thread.

@@

A CPU may start executing an instruction speculatively, but at some
point it must either abandon that instruction or \emph{commit} it with
a set of specific outputs based on well-defined inputs.
A CPU is clearly not permitted to commit an instruction its inputs have
been committed, unless the remaining uncommitted inputs cannot affect
that instruction's outputs.
For instance, a conditional-move instruction needn't wait for the source
input to commit if it knows that the condition is definitely false.

The overall effect of these hardware dependencies is that if a change to an
output of instruction $I$ would lead to a change in the action or outputs of
a later instruction $J$, then the CPU must commit $I$'s output before
committing $J$'s action and outputs.
Thus dependencies force the affected instructions to commit in order,
even on weakly ordered architectures.
We will say that one instruction is \emph{ordered after} another to mean that
it must commit after the other one commits.\footnote{
	Note that this implies nothing about when a load instruction
	retrieves its value from memory; it may do so long before it
	commits.}

Dependencies aren't the only ordering mechanism.
As mentioned in Section~\ref{sec:Brief OOTA Overview}, a CPU does
not make the value of a store instruction available to other CPUs
until after the store commits.
It then takes additional time for the stored value to travel to other
CPUs, owing to finite speed of light and non-zero size of computing
hardware.
Since a load instruction cannot commit until its output (the value
read) is fully determined, it must commit after the store it reads
from.

Conditional or computed branches also provide ordering.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to;
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional- or computed-branch
instruction must commit after the branch, and hence after the source
for the branch's condition and/or destination
input.\footnote{
	One exception is a branch that conditionally jumps to the
	next instruction.
	We ignore this possibility by treating such branches as no-ops.}
